---
output:
  pdf_document:
    number_sections: true
    # template: tex/template.tex
bibliography: tex/bibliography.bib
biblio-style: text/econometrica.bst
header-includes:
  - \usepackage{floatrow}
  - \floatsetup{capposition=top}
  - \usepackage{setspace}
editor_options: 
  markdown: 
    wrap: 72
---

<!-- Setup -------------------------------------------------------------------->

```{r setup, include=FALSE}
if(!require(kableExtra)) install.packages("kableExtra")
if(!require(randomForest)) install.packages("randomForest")
if(!require(iml)) install.packages("iml")

library(kableExtra)
library(randomForest)
library(iml)

knitr::opts_chunk$set(echo = FALSE, fig.align = "center")

load("data/input.RData")
varnames <- list.export[["varnames"]]
```

<!-- Title Page + TOC --------------------------------------------------------->

```{=tex}
\newpage
\vspace*{4cm}
\begin{center}
{\Large Machine Learning in Economics \\ (458657)}
\end{center}
\vspace*{1.5cm}
\begin{center}
\thispagestyle{empty}
{\LARGE \bf replication paper}\\[1.5cm]
{\Huge Early Warning System for Fiscal Stress}\\[1.5cm]
{\bf \Large comparing the traditional logistic regression approach \\ with random forest}\\[2cm]
{\bf \large Department of Economics\\
University of Bern}\\[1cm]
{\Large Spring Semester 2022}\\[1.5cm]
{\large submitted by Bela Tim Koch} \\[2.5cm]
\end{center}
```
```{=tex}
\newpage
\thispagestyle{empty}
\tableofcontents
\newpage
```

<!-- Report ------------------------------------------------------------------->
<!--  1 Title Page
      2 TOC
      3 Introduction + Literature Review
      4-5.5 Model Description
      5.5-7 Data Description
      8-9 Results
      10 Conclusion
-->


\setcounter{page}{1}
\onehalfspace

# Introduction and Literature Review

Since the latest Great Recession with its corresponding deterioration of public
finances, the monitoring and prevention of fiscal crises has become increasingly
prominent in the political debate, leading to an increasing demand for the
development of reliable and early indicators that signal possible fiscal stress.
In order to be able to assess countries' vulnerability to fiscal distress ex-ante,
the literature is increasingly devoted to the development of early warning
systems for fiscal stress, which build on early warning systems for banking and
currency crises [@honda2022]. The standard tool used in the literature for early
warning systems are the signalling approach as well as discrete dependent variable
models, such as logistic regression [@jarmulska2020]. However, as an alternative
to the traditional methods, early warning models based on machine learning
techniques are proposed, claiming a possible improvement of prediction accuracy
[@beutel2019].

<!-- HIER EIN PAAR BEISPIEL PAPERS AUFFÜHREN. -->

This paper aims to replicate some of the work done by @jarmulska2020. Particular
emphasis is placed on the comparison of the traditional method, i.e. a logit model
with a least absolute shrinkage and selection operator, and a novel approach of
using an implementation of a random forest algorithm. Since machine learning
algorithms such as the random forest is often accused of a lack of interpretability,
ways of interpreting the developed model are presented.


\newpage

# Model Describtion

## Performance Metrics

@jarmulska2020 uses sensitivity, specificity, their average as well as
the area under receiver operating curve (AUROC) as measures to assess
the effectiveness of the early warning models. Since sensitivity
corresponds to the proportion of stress episodes correctly classified
whereby specificity corresponds to the proportion of tranquil episodes
correctly classified, these metrics are dependent on the threshold,
which determines whether a period is classified as stress or tranquil
episode [@jarmulska2020]. In this paper, this threshold is specified by
maximizing the weighted sum of sensitivity and specificity. In contrast,
the AUROC is a robust measure, as all possible thresholds are taken into
account in their calculation. This measure represents the area under the
receiver operating curve, which displays the trade-off between the true
positive rate (i.e. sensitivity) and the false positive rate (i.e. 1 -
specificity). Theoretically, the AUROC can be between 0 and 1 (perfect
classifier), whereby random guessing would result in to a value of 0.5
[@fawcett2006].

## Logit Model with LASSO penalisation

@jarmulska2020 implemented two version of discrete dependent variable
models (logit regression), first a standard logit model with ordinary
least squares estimates and second a logit model with a least absolute
shrinkage and selection operator (LASSO) penalization. These models are
often used as the standard econometric approach, which is why they are
used as the benchmark in this study.

Ordinary least squares estimates often have low bias but large variance,
reducing prediction accuracy - the prediction accuracy can sometimes be
improved by shrinking some coefficients towards zero to sacrifice bias
in order to reduce variance of the predicted values and possibly improve
overall prediction accuracy [@tibshirani1996]. To do so, LASSO
penalization as proposed in @tibshirani1996 can be applied. Because of
this characteristics, only the logit LASSO model is considered in this
replication.

Following @hastie2009, the LASSO problem in the Lagrangian form is given
as follows:

```{=tex}
\begin{equation} \label{eq:lasso}
\hat{\beta}^{lasso} = \underset{\beta}{\text{argmin}} \left\{
\frac{1}{2} \sum^N_{i=1}(y_i - \beta_0 - \sum^p_{j=1}x_{ij}\beta_j)^2
+ \lambda \sum^p_{j=1}|\beta_j| \right\}
\end{equation}
```
whereby $\lambda$ corresponds to the penalization parameter. As can be
seen in Equation \eqref{eq:lasso}, the higher $\lambda$, the higher the
number of coefficients shrunk to zero. Here, $\lambda$ is chosen by
5-fold cross-validation, maximizing the AUROC.

## Random Forest

<!-- As a second method of building an early warning system, @jarmulska2020 applied -->

<!-- classification and regression trees (CART) for binary classification and their -->

<!-- ensemble into random forests.  -->

Building upon decision trees, which divide the predictor space into

zuerst ganz kurz definition cart (ev. mit anderer quelle als jarmulska)

dann ganz kurs warum cart zu rf

dann ganz kurz rf

Gini index

```{=tex}
\begin{equation}
g(w) = \sum_{k \neq j}p_{wk}p_{wj} = \sum_k p_{wk}(1-p_{wk})
\end{equation}
```
\newpage

# Data Describtion

## Dependent Variable

definition of a fiscal stress event follows @dobrescu2011 siehe die
tabelle

empirical/historical data about fiscal stress events

```{r stress.distr, fig.height=3, fig.cap="Distribution of Stress Periods"}
data <- list.export[["data"]]
dt <- data


dt$year <- dt$year+2
dt$crisis_next_period <- as.numeric(as.character(dt$crisis_next_period))


dt_developing <- dt[dt$developed == 0,]
dt_developed  <- dt[dt$developed == 1,]

dt_developing <- dt_developing[, names(dt_developing) %in% c("year", "crisis_next_period")]
dt_developing <- aggregate(dt_developing$crisis_next_period,
                           by = list(dt_developing$year),
                           FUN = sum)
colnames(dt_developing) <- c("year", "sum_developing")

dt_developed <- dt_developed[, names(dt_developed) %in% c("year", "crisis_next_period")]
dt_developed <- aggregate(dt_developed$crisis_next_period,
                           by = list(dt_developed$year),
                           FUN = sum)

colnames(dt_developed) <- c("year", "sum_developed")

dt <- merge(x = dt_developing, y = dt_developed,
            by = "year",
            all = T)

par(las = 1, cex = 0.6)
barplot(rbind(dt$sum_developing, dt$sum_developed), beside = T,
        names.arg = dt$year, las = 2, col = c("gray30", "gray90"))
legend("topleft",
       legend = c("developing countries", "developed countries"),
       fill = c("gray30", "gray90"))


```

## Explanatory Variables

```{r means.table}
# load relevant data
means.table <- list.export[["means.table"]]

# round numeric columns
num.cols <- sapply(means.table, mode) == "numeric"
means.table[num.cols] <- round(means.table[num.cols], 2)

# define categories
macro_globecon <- c("interest_rate_US", "dyn_GDP_US", "dyn_gdp_china", "oil_yoy",
                    "VIX", "dyn_gdp", "GDP_per_cap")
comp_domdem <- c("overvaluation", "ca_balance", "dyn_export_share", "dyn_fix_cap_form",
                 "cpi", "dyn_consum")
fin <- c("dyn_fx_rate", "diff_priv_credit_gdp")
fisc <- c("net_lending", "public_debt", "interest_on_debt")
labor <- c("diff_unempl", "dyn_prod_dol")

# add categories
means.table$category <- NA
means.table$category <- ifelse(means.table$variable %in% macro_globecon,
                               "Macroeconomic and global economy",
                               means.table$category)
means.table$category <- ifelse(means.table$variable %in% comp_domdem,
                               "Competitiveness and domestic demand",
                               means.table$category)
means.table$category <- ifelse(means.table$variable %in% fin,
                               "Financial",
                               means.table$category)
means.table$category <- ifelse(means.table$variable %in% fisc,
                               "Fiscal",
                               means.table$category)
means.table$category <- ifelse(means.table$variable %in% labor,
                               "Labor market",
                               means.table$category)

# add variable names
means.table <- merge(x = varnames, y = means.table,
                     by = "variable",
                     all = T)
means.table$variable <- NULL

# recode significance
means.table$significant <- ifelse(means.table$significant == T, "yes", "no")

# rename columns
names(means.table)[names(means.table) == "name"] <- "Variable"
names(means.table)[names(means.table) == "all_periods"] <- "All periods"
names(means.table)[names(means.table) == "tranq_periods"] <- "Tranquil periods"
names(means.table)[names(means.table) == "stress_periods"] <- "Stress periods"
names(means.table)[names(means.table) == "p_value"] <- "P-value"
names(means.table)[names(means.table) == "significant"] <- "Significance"

# make table
kbl(means.table[, -which(names(means.table) == "category")], booktabs = T,
    caption = "Means of Explanatory Variables", align = c("l", rep("r", 5))) %>%
  pack_rows(index = table(means.table$category)) %>%
  kable_styling(latex_options="scale_down")
```

\newpage

# Empirical Results

## Performance

```{r avg-pred-accur}
# get relevant data
dt <- list.export[["results.avg"]]

# manipulate formation
dt[!colnames(dt) %in% c("model", "auc")] <- apply(dt[!colnames(dt) %in% c("model", "auc")], 2, function(x) x*100)
dt[colnames(dt) != "model"] <- apply(dt[colnames(dt) != "model"], 2, function(x) round(x, 2))
```

```{=tex}
\begin{table}[H]
\centering
\begin{tabular}{@{\extracolsep{4pt}}lrrrr@{}}
                                             & \multicolumn{2}{c}{Logit LASSO}                                         & \multicolumn{2}{c}{Random Forest}                                       \\ \cline{2-3} \cline{4-5}
                                             & \multicolumn{1}{c}{\begin{tabular}[c]{@{}r@{}}advanced\\ dummy\end{tabular}} & \multicolumn{1}{c}{\begin{tabular}[c]{@{}r@{}}GDP\\ per capita\end{tabular}} & \multicolumn{1}{c}{\begin{tabular}[c]{@{}r@{}}advanced\\ dummy\end{tabular}} & \multicolumn{1}{c}{\begin{tabular}[c]{@{}r@{}}GDP\\ per capita\end{tabular}} \\ \hline
\begin{tabular}[c]{@{}l@{}}\% of correctly\\ classified stress episodes\end{tabular}   & `r dt[dt$model == "logit.lasso.DUMMY",]$prop.pos` & `r dt[dt$model == "logit.lasso.GDP",]$prop.pos` & `r dt[dt$model == "rf.DUMMY",]$prop.pos` & `r dt[dt$model == "rf.GDP",]$prop.pos` \\ \hline
\begin{tabular}[c]{@{}l@{}}\% of correctly\\ classified tranquil episodes\end{tabular} & `r dt[dt$model == "logit.lasso.DUMMY",]$prop.neg` & `r dt[dt$model == "logit.lasso.GDP",]$prop.neg` & `r dt[dt$model == "rf.DUMMY",]$prop.neg` & `r dt[dt$model == "rf.GDP",]$prop.neg` \\ \hline
\begin{tabular}[c]{@{}l@{}}Average\\ \phantom{Average} \end{tabular}                                      & `r dt[dt$model == "logit.lasso.DUMMY",]$avg`      & `r dt[dt$model == "logit.lasso.GDP",]$avg`      & `r dt[dt$model == "rf.DUMMY",]$avg`      & `r dt[dt$model == "rf.GDP",]$avg`      \\ \hline
\begin{tabular}[c]{@{}l@{}}AUROC\\ \phantom{AUROC} \end{tabular}                                        & `r dt[dt$model == "logit.lasso.DUMMY",]$auc`      & `r dt[dt$model == "logit.lasso.GDP",]$auc`      & `r dt[dt$model == "rf.DUMMY",]$auc`      & `r dt[dt$model == "rf.GDP",]$auc`      \\ \hline
\end{tabular}
\caption{\label{tab:avg-pred-accur}Average prediction accuracy of early warning models for years 2009-2018}
\end{table}
```
## Interpretability of Random Forest Algorithm

### Variable Importance and Shapley Values

```{r rf.VI.SV, fig.cap="Variable Importance and Shapley Values of Predictors used", fig.height=3}
# prepare data for plot for variable importance
rf.fit.eval <- list.export[["rf.fit.eval"]]

varimp <- as.data.frame(importance(rf.fit.eval))
varimp$variable <- rownames(varimp)
varimp <- varimp[order(varimp$MeanDecreaseGini, decreasing = T), c(1,2)]
varimp$yaxis <- rev(1:nrow(varimp))

varimp <- merge(x = varimp, y = varnames,
                by = "variable",
                all = T)

varimp <- varimp[order(varimp$MeanDecreaseGini, decreasing = T), c(1:4)]
labs.varimp <- rev(varimp$name)

# prepare data for plot for shapley values
shapley.values <- list.export[["shapley.values"]]

shapley.values <- shapley.values[order(shapley.values$x, decreasing = T), c(1,2)]
shapley.values$yaxis <- rev(1:nrow(shapley.values))

shapley.values <- merge(x = shapley.values, y = varnames,
                        by.x = "Group.1", by.y = "variable",
                        all = T)

shapley.values <- shapley.values[order(shapley.values$x, decreasing = T), c(1:4)]
labs.shapley <- rev(shapley.values$name)

# draw plot
par(mar = c(5.1, 10.5, 2.1, 0.5), mfrow = c(1,2), las=1, cex = 0.75, cex.axis = 0.75, mgp = c(1.75, 0.75, 0))

# plot 1: variable importance
plot(x = varimp$MeanDecreaseGini, y = varimp$yaxis,
     main = "", xlab = "Breiman's Variable Importance", ylab = "", yaxt = "none",
     pch = 16)
abline(h = 1:20, lty = 3)
axis(side = 2, at = 1:20, labels = labs.varimp)

# plot 2: shapley values
plot(x = shapley.values$x, y = shapley.values$yaxis,
     main = "", xlab = "Shapley Values", ylab = "", yaxt = "none", pch = 16)
abline(h = 1:20, lty = 3)
axis(side = 2, at = 1:20, labels = labs.shapley)
```

### Partial dependence plots and Accumulated local effects plots

```{r rf.pdp.alep, fig.cap="Partial Dependence and Accumulated Local Effects Plots"}

# load data for partial dependence plots
partial.ca_balance  <- list.export[["partial.ca_balance"]]
partial.net_lending <- list.export[["partial.net_lending"]]

# load data for accumulated local effects plots
ale.ca_balance  <- list.export[["ale.ca_balance"]]
ale.net_lending <- list.export[["ale.net_lending"]]

# prepare plot
# x.train.eval.df <- list.export[["x.train.eval.df"]]
x.train.eval.df <- as.data.frame(list.export[["x.train.eval"]])
par(mfrow = c(2,2), mar = c(5.1, 4.1, 1, 2.1), cex = 0.65, cex.axis = 0.75, las = 1, mgp = c(2.25, 0.55, 0))

# plot upper left: pdp for ca balance
plot(x = partial.ca_balance$ca_balance, y = partial.ca_balance$yhat,
           type = "l", xlab = "Current account balance",
           ylab = "Predicted probability of stress")
rug(x.train.eval.df$ca_balance, ticksize = 0.015)

# plot upper right: ace plot for ca balance
plot(x = ale.ca_balance$results$ca_balance, y = ale.ca_balance$results$.value,
           type = "l", xlab = "Current account balance",
           ylab = "Predicted probability of stress")
rug(x.train.eval.df$ca_balance, ticksize = 0.015)

# plot lower left: pdp for net lending
plot(x = partial.net_lending$net_lending, y = partial.net_lending$yhat,
           type = "l", xlab = "Net lending",
           ylab = "Predicted probability of stress")
rug(x.train.eval.df$net_lending, ticksize = 0.015)

# plot lower right: ace plot net lending 
plot(x = ale.net_lending$results$net_lending, y = ale.net_lending$results$.value,
           type = "l", xlab = "Net lending",
           ylab = "Predicted probability of stress")
rug(x.train.eval.df$net_lending, ticksize = 0.015)
```

\newpage

# Conclusion

<!-- References --------------------------------------------------------------->

\newpage

# References

The code and data used for this project can be found in the
corresponding GitHub-repository:
\mbox{\texttt{\url{https://github.com/bt-koch/ML-in-Economics}}}.

\vspace{1cm}

\bibliography{tex/bibliography.bib}
